---
title: "CMHEF vs CMDEF exp 3-6, old"
output:
  workflowr::wflow_html:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    code_folding: hide
---

# Setup
```{r, include=FALSE}
#load packages
# Statistical modelling
library(MASS)
library(MuMIn)
library(broom)

# Core data manipulation and plotting
# Loaded after MASS so dplyr functions take precedence
library(tidyverse)

# Plotting and figure assembly
library(pheatmap)
library(RColorBrewer)
library(UpSetR)
library(gridExtra)

# Differential expression and Bioconductor tools
library(limma)
library(scDblFinder)
library(rhdf5)
library(here)

# Single-cell transcriptomics
library(Seurat)
library(sctransform)
library(scCustomize)
library(harmony)
library(clustree)
library(presto)

set.seed(123)
```
# Read in and prepare Datasets
```{r }
# Sample 3 healthy
Exp3_CMHEF_NC <- here("data",  "cellbender_output_Exp3_CMHEF_NC_filtered.h5")
counts_3H <- Read_CellBender_h5_Mat(file_name = Exp3_CMHEF_NC)
preQC_3H <- CreateSeuratObject(counts = counts_3H,  project = "Exp3_CMHEF_NC")

# Sample 3 diseased
Exp3_CMDEF_NC <- here("data", "cellbender_output_Exp3_CMDEF_NC_filtered.h5")
counts_3D <- Read_CellBender_h5_Mat(file_name = Exp3_CMDEF_NC)
preQC_3D <- CreateSeuratObject(counts = counts_3D, project = "Exp3_CMDEF_NC")

# Sample 4 healthy 
Exp4_CMHEF_NC <- here("data","cellbender_output_Exp4_CMHEF_NC_filtered.h5")
counts_4H <- Read_CellBender_h5_Mat(file_name = Exp4_CMHEF_NC)
preQC_4H <- CreateSeuratObject(counts = counts_4H, project = "Exp4_CMHEF_NC")

# Sample 4 diseased 
Exp4_CMDEF_NC <- here("data","cellbender_output_Exp4_CMDEF_NC_filtered.h5")
counts_4D <- Read_CellBender_h5_Mat(file_name = Exp4_CMDEF_NC)
preQC_4D <- CreateSeuratObject(counts = counts_4D,project = "Exp4_CMDEF_NC")

#sample 5 healthy
Exp5_CMHEF_NC <- here("data","cellbender_output_Exp5_CMHEF_NC_filtered.h5")
counts_5H <- Read_CellBender_h5_Mat(file_name = Exp5_CMHEF_NC)
preQC_5H <- CreateSeuratObject(counts = counts_5H,project = "Exp5_CMHEF_NC")

#sample 5 diseased
Exp5_CMDEF_NC <- here("data","cellbender_output_Exp5_CMDEF_NC_filtered.h5")
counts_5D <- Read_CellBender_h5_Mat(file_name = Exp5_CMDEF_NC)
preQC_5D <- CreateSeuratObject(counts = counts_5D,project = "Exp5_CMDEF_NC")

#sample 6 healthy
Exp6_CMHEF_NC <- here("data","cellbender_output_Exp6_CMHEF_NC_filtered.h5")
counts_6H <- Read_CellBender_h5_Mat(file_name = Exp6_CMHEF_NC)
preQC_6H <- CreateSeuratObject(counts = counts_6H,project = "Exp6_CMHEF_NC")

#sample 6 diseased
Exp6_CMDEF_NC <- here("data","cellbender_output_Exp6_CMDEF_NC_filtered.h5")
counts_6D <- Read_CellBender_h5_Mat(file_name = Exp6_CMDEF_NC)
preQC_6D <- CreateSeuratObject(counts = counts_6D,project = "Exp6_CMDEF_NC")
```

```{r }
# Sample and condition metadata

preQC_3H$sample <- "Exp3_CMHEF_NC"
preQC_3H$condition <- "healthy"

preQC_3D$sample <- "Exp3_CMDEF_NC"
preQC_3D$condition <- "diseased"

preQC_4H$sample <- "Exp4_CMHEF_NC"
preQC_4H$condition <- "healthy"

preQC_4D$sample <- "Exp4_CMDEF_NC"
preQC_4D$condition <- "diseased"

preQC_5H$sample <- "Exp5_CMHEF_NC"
preQC_5H$condition <- "healthy"

preQC_5D$sample <- "Exp5_CMDEF_NC"
preQC_5D$condition <- "diseased"

preQC_6H$sample <- "Exp6_CMHEF_NC"
preQC_6H$condition <- "healthy"

preQC_6D$sample <- "Exp6_CMDEF_NC"
preQC_6D$condition <- "diseased"


# Biological replicate metadata

preQC_3H$replicate <- "Exp3"
preQC_3D$replicate <- "Exp3"

preQC_4H$replicate <- "Exp4"
preQC_4D$replicate <- "Exp4"

preQC_5H$replicate <- "Exp5"
preQC_5D$replicate <- "Exp5"

preQC_6H$replicate <- "Exp6"
preQC_6D$replicate <- "Exp6"

```

# PreQC Plots
```{r, fig.width=8, fig.height=5, fig.align='center'}

# Add mitochondrial percentage to all samples
preQC_3H[["percent.mt"]] <- PercentageFeatureSet(preQC_3H, pattern = "^MT-")
preQC_3D[["percent.mt"]] <- PercentageFeatureSet(preQC_3D, pattern = "^MT-")
preQC_4H[["percent.mt"]] <- PercentageFeatureSet(preQC_4H, pattern = "^MT-")
preQC_4D[["percent.mt"]] <- PercentageFeatureSet(preQC_4D, pattern = "^MT-")
preQC_5H[["percent.mt"]] <- PercentageFeatureSet(preQC_5H, pattern = "^MT-")
preQC_5D[["percent.mt"]] <- PercentageFeatureSet(preQC_5D, pattern = "^MT-")
preQC_6H[["percent.mt"]] <- PercentageFeatureSet(preQC_6H, pattern = "^MT-")
preQC_6D[["percent.mt"]] <- PercentageFeatureSet(preQC_6D, pattern = "^MT-")


# Add ribosomal protein percentage to all samples
preQC_3H[["percent.rp"]] <- PercentageFeatureSet(preQC_3H, pattern = "^RP[SL]")
preQC_3D[["percent.rp"]] <- PercentageFeatureSet(preQC_3D, pattern = "^RP[SL]")
preQC_4H[["percent.rp"]] <- PercentageFeatureSet(preQC_4H, pattern = "^RP[SL]")
preQC_4D[["percent.rp"]] <- PercentageFeatureSet(preQC_4D, pattern = "^RP[SL]")
preQC_5H[["percent.rp"]] <- PercentageFeatureSet(preQC_5H, pattern = "^RP[SL]")
preQC_5D[["percent.rp"]] <- PercentageFeatureSet(preQC_5D, pattern = "^RP[SL]")
preQC_6H[["percent.rp"]] <- PercentageFeatureSet(preQC_6H, pattern = "^RP[SL]")
preQC_6D[["percent.rp"]] <- PercentageFeatureSet(preQC_6D, pattern = "^RP[SL]")



# Merge all samples for QC visualization
qc_combined <- merge(
  preQC_3H,
  y = list(
    preQC_3D,
    preQC_4H,
    preQC_4D,
    preQC_5H,
    preQC_5D,
    preQC_6H,
    preQC_6D
  ),
  add.cell.ids = c(
    "Exp3H",
    "Exp3D",
    "Exp4H",
    "Exp4D",
    "Exp5H",
    "Exp5D",
    "Exp6H",
    "Exp6D"
  ),
  project = "QC_compare"
)


# Add abbreviated sample names
qc_combined$sample_short <- dplyr::recode(
  qc_combined$sample,
  "Exp3_CMHEF_NC" = "3H",
  "Exp3_CMDEF_NC" = "3D",
  "Exp4_CMHEF_NC" = "4H",
  "Exp4_CMDEF_NC" = "4D",
  "Exp5_CMHEF_NC" = "5H",
  "Exp5_CMDEF_NC" = "5D",
  "Exp6_CMHEF_NC" = "6H",
  "Exp6_CMDEF_NC" = "6D"
)


# Plot number of detected genes per nucleus
VlnPlot(
  qc_combined,
  features = "nFeature_RNA",
  group.by = "sample",
  pt.size = 0,
  ncol = 3
) +
  NoLegend() +
  ggtitle("nFeature_RNA") +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    plot.margin = margin(
      t = 5,
      r = 5,
      b = 50,
      l = 5
    )
  )


# Plot total RNA counts per nucleus
VlnPlot(
  qc_combined,
  features = "nCount_RNA",
  group.by = "sample",
  pt.size = 0,
  ncol = 3
) +
  NoLegend() +
  ggtitle("nCount_RNA") +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    plot.margin = margin(
      t = 5,
      r = 5,
      b = 50,
      l = 5
    )
  )


# Plot mitochondrial percentage
VlnPlot(
  qc_combined,
  features = "percent.mt",
  group.by = "sample",
  pt.size = 0,
  ncol = 3
) +
  NoLegend() +
  ggtitle("percent.mt") +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    plot.margin = margin(
      t = 10,
      r = 10,
      b = 20,
      l = 40
    )
  )


# Plot ribosomal protein percentage
VlnPlot(
  qc_combined,
  features = "percent.rp",
  group.by = "sample",
  pt.size = 0,
  ncol = 3
) +
  NoLegend() +
  ggtitle("percent.rp") +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    plot.margin = margin(
      t = 10,
      r = 10,
      b = 20,
      l = 40
    )
  )
```



# Doublet Removal
```{r fig.width=14, fig.height=10, dpi=300}
# Doublet removal 

sample_list <- list(
  Exp3_CMHEF_NC = preQC_3H,
  Exp3_CMDEF_NC = preQC_3D,
  Exp4_CMHEF_NC = preQC_4H,
  Exp4_CMDEF_NC = preQC_4D,
  Exp5_CMHEF_NC = preQC_5H,
  Exp5_CMDEF_NC = preQC_5D,
  Exp6_CMHEF_NC = preQC_6H,
  Exp6_CMDEF_NC = preQC_6D
)


# Run scDblFinder separately for each sample
for (nm in names(sample_list)) {
  
  sce_obj <- as.SingleCellExperiment(sample_list[[nm]])
  
  set.seed(123)
  
  sce_obj <- scDblFinder(
    sce_obj,
    clusters = TRUE
  )
  
  sample_list[[nm]]$scDblFinder.class <-
    sce_obj$scDblFinder.class
  
  sample_list[[nm]]$scDblFinder.score <-
    sce_obj$scDblFinder.score
}


# Merge all samples before doublet removal
preQC_combined <- merge(
  x = sample_list[[1]],
  y = sample_list[-1],
  add.cell.ids = names(sample_list),
  project = "preQC_combined"
)


# Set the order of doublet classifications
preQC_combined$scDblFinder.class <- factor(
  preQC_combined$scDblFinder.class,
  levels = c("singlet", "doublet")
)


# Compare QC metrics between singlets and doublets
VlnPlot(
  preQC_combined,
  features = c(
    "nCount_RNA",
    "nFeature_RNA",
    "percent.mt"
  ),
  group.by = "sample",
  split.by = "scDblFinder.class",
  layer = "counts",
  pt.size = 0,
  ncol = 3
) & theme(
  legend.position = "right",
  axis.text.x = element_text(
    angle = 45,
    hjust = 1
  )
)


# Retain only nuclei classified as singlets
sample_list_nodoublets <- lapply(
  sample_list,
  function(x) {
    subset(
      x = x,
      subset = scDblFinder.class == "singlet"
    )
  }
)


# Merge all samples after doublet removal
preQC_nodoublets_combined <- merge(
  x = sample_list_nodoublets[[1]],
  y = sample_list_nodoublets[-1],
  add.cell.ids = names(sample_list_nodoublets),
  project = "preQC_nodoublets_combined"
)


# Visualize QC metrics after doublet removal
VlnPlot(
  preQC_nodoublets_combined,
  features = c(
    "nFeature_RNA",
    "nCount_RNA",
    "percent.mt"
  ),
  group.by = "sample",
  layer = "counts",
  pt.size = 0,
  ncol = 3
) & theme(
  axis.text.x = element_text(
    angle = 45,
    hjust = 1
  )
)


# Display doublet classifications for each sample
lapply(
  sample_list,
  function(x) {
    table(x$scDblFinder.class)
  }
)


# Create a summary table of singlets and doublets
doublet_table <- do.call(
  rbind,
  lapply(
    names(sample_list),
    function(nm) {
      
      tab <- table(
        sample_list[[nm]]$scDblFinder.class
      )
      
      singlet_count <- if (
        "singlet" %in% names(tab)
      ) {
        as.numeric(tab["singlet"])
      } else {
        0
      }
      
      doublet_count <- if (
        "doublet" %in% names(tab)
      ) {
        as.numeric(tab["doublet"])
      } else {
        0
      }
      
      total_count <- sum(tab)
      
      data.frame(
        Sample = nm,
        Singlet = singlet_count,
        Doublet = doublet_count,
        Total = total_count,
        Doublet_percent = round(
          100 * doublet_count / total_count,
          1
        )
      )
    }
  )
)


# View doublet summary
doublet_table
```


```{r }
#some checks after doublet removal to be sure that the seurat objects are correct
lapply(sample_list, function(x) table(x$scDblFinder.class))
lapply(sample_list_nodoublets, function(x) table(x$scDblFinder.class))

sapply(sample_list, ncol)
sapply(sample_list_nodoublets, ncol)


```

# PostQC plots
```{r fig.width=14, fig.height=10, dpi=300}

#use QC parameters as when I looked at the samples on their own as I did before
postQC_3H <- subset(sample_list_nodoublets$Exp3_CMHEF_NC, subset = nFeature_RNA > 600 & percent.mt < 3)
postQC_3D <- subset(sample_list_nodoublets$Exp3_CMDEF_NC, subset = nFeature_RNA > 800 & percent.mt < 3)
postQC_4H <- subset(sample_list_nodoublets$Exp4_CMHEF_NC, subset = nFeature_RNA > 800 & percent.mt < 3)
postQC_4D <- subset(sample_list_nodoublets$Exp4_CMDEF_NC, subset = nFeature_RNA > 900 & percent.mt < 3)
postQC_5H <- subset(sample_list_nodoublets$Exp5_CMHEF_NC, subset = nFeature_RNA > 800 & percent.mt < 3)
postQC_5D <- subset(sample_list_nodoublets$Exp5_CMDEF_NC, subset = nFeature_RNA > 800 & percent.mt < 3)
postQC_6H <- subset(sample_list_nodoublets$Exp6_CMHEF_NC, subset = nFeature_RNA > 800 & percent.mt < 3)
postQC_6D <- subset(sample_list_nodoublets$Exp6_CMDEF_NC, subset = nFeature_RNA > 800 & percent.mt < 3)

Post_QC_plot <- merge(
  postQC_3H,
  y = list(postQC_3D, postQC_4H, postQC_4D, postQC_5H, postQC_5D, postQC_6H, postQC_6D),
  add.cell.ids = c("Exp3H", "Exp3D", "Exp4H", "Exp4D", "Exp5H", "Exp5D", "Exp6H", "Exp6D"),
  project = "Post_QC"
)

VlnPlot(
  Post_QC_plot,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rp"),
  group.by = "sample",
  ncol = 4,
  pt.size = 0
) & theme(
  axis.text.x = element_text(angle = 45, hjust = 1)
)
```

# Normalization
```{r }

#normalize, each sample individually, and new seed
set.seed(124)
# normalization (per sample, before Harmony)

options(future.globals.maxSize = 2 * 1024^3)  # 2 GB

Norm_postQC_3H <- SCTransform(
  postQC_3H,
  assay = "RNA",
  vst.flavor = "v2",
  vars.to.regress = "percent.mt"
)

Norm_postQC_3D <- SCTransform(
  postQC_3D,
  assay = "RNA",
  vst.flavor = "v2",
  vars.to.regress = "percent.mt"
)

Norm_postQC_4H <- SCTransform(
  postQC_4H,
  assay = "RNA",
  vst.flavor = "v2",
  vars.to.regress = "percent.mt"
)

Norm_postQC_4D <- SCTransform(
  postQC_4D,
  assay = "RNA",
  vst.flavor = "v2",
  vars.to.regress = "percent.mt"
)

Norm_postQC_5H <- SCTransform(
  postQC_5H,
  assay = "RNA",
  vst.flavor = "v2",
  vars.to.regress = "percent.mt"
)

Norm_postQC_5D <- SCTransform(
  postQC_5D,
  assay = "RNA",
  vst.flavor = "v2",
  vars.to.regress = "percent.mt"
)

Norm_postQC_6H <- SCTransform(
  postQC_6H,
  assay = "RNA",
  vst.flavor = "v2",
  vars.to.regress = "percent.mt"
)

Norm_postQC_6D <- SCTransform(
  postQC_6D,
  assay = "RNA",
  vst.flavor = "v2",
  vars.to.regress = "percent.mt"
)


# Merge the SCT-normalized objects
Norm_Post_QC <- merge(
  x = Norm_postQC_3H,
  y = list(
    Norm_postQC_3D,
    Norm_postQC_4H,
    Norm_postQC_4D,
    Norm_postQC_5H,
    Norm_postQC_5D,
    Norm_postQC_6H,
    Norm_postQC_6D
  ),
  add.cell.ids = c("3H", "3D", "4H", "4D", "5H", "5D", "6H", "6D"),
  project = "Post_QC"
)

# Use SCT assay for downstream PCA
DefaultAssay(Norm_Post_QC) <- "SCT"
```



# PCA
```{r }
#linear dimensionality reduction
set.seed(125)
# choose one shared variable-feature set across the 8 SCT-normalized samples
features <- SelectIntegrationFeatures(
  object.list = list(
    Norm_postQC_3H,
    Norm_postQC_3D,
    Norm_postQC_4H,
    Norm_postQC_4D,
    Norm_postQC_5H,
    Norm_postQC_5D,
    Norm_postQC_6H,
    Norm_postQC_6D
  ),
  nfeatures = 3000
)

# assign those shared features to the merged object
VariableFeatures(Norm_Post_QC) <- features

# make sure PCA uses the SCT assay
DefaultAssay(Norm_Post_QC) <- "SCT"

# now run PCA
Norm_Post_QC <- RunPCA(
  Norm_Post_QC,
  features = VariableFeatures(Norm_Post_QC)
)
```

# Visualize PCA
```{r }
#visualizations of the PCA
VizDimLoadings(Norm_Post_QC, dims = 1:2, reduction = "pca")

DimPlot(Norm_Post_QC, reduction = "pca") + NoLegend()

DimHeatmap(Norm_Post_QC, dims = 1, cells = 500, balanced = TRUE)

DimHeatmap(Norm_Post_QC, dims = 1:15, cells = 500, balanced = TRUE)

ElbowPlot(Norm_Post_QC)

ElbowPlot(Norm_Post_QC, ndims = 50) +
  scale_y_continuous(breaks = seq(0, 50, by = 2))

```

# pre-Harmony UMAP
```{r }
set.seed(126)
#before integration
# pre-Harmony UMAP
Norm_Post_QC <- RunUMAP(
  Norm_Post_QC,
  reduction = "pca",
  dims = 1:40,
  reduction.name = "umap.pca"
)

DimPlot(Norm_Post_QC, reduction = "umap.pca", group.by = "sample")
DimPlot(Norm_Post_QC, reduction = "umap.pca", group.by = "condition")



```
# post-Harmony UMAP
```{r }
#do harmony and after harmony umap
set.seed(127)
Norm_Post_QC <- RunHarmony(
  object = Norm_Post_QC,
  group.by.vars = "sample",
  reduction.use = "pca",
  dims.use=1:40,
  plot_convergence = TRUE
)

# post-Harmony UMAP
Norm_Post_QC <- RunUMAP(
  Norm_Post_QC,
  reduction = "harmony",
  dims = 1:40,
  reduction.name = "umap.harmony"
)

DimPlot(Norm_Post_QC, reduction = "umap.harmony", group.by = "sample")
DimPlot(Norm_Post_QC, reduction = "umap.harmony", group.by = "condition")



```

# Clustree
```{r fig.width=12, fig.height=9, dpi=300}
set.seed(128)
# clustering on Harmony
Norm_Post_QC <- FindNeighbors(
  Norm_Post_QC,
  reduction = "harmony",
  dims = 1:40
)


#figure out which resolution I want/need
Norm_Post_QC <- FindClusters(Norm_Post_QC, resolution = seq(0.1, 0.8, by=0.1))
clustree::clustree(Norm_Post_QC@meta.data[,grep("SCT_snn_res", colnames(Norm_Post_QC@meta.data))],
                   prefix = "SCT_snn_res.")

```

# Chose a resolution for UMAP
```{r }
set.seed(129)
Norm_Post_QC_0.3 <- FindClusters(
  Norm_Post_QC,
  resolution = 0.3
)
```


# UMAP Resolution 0.3
```{r fig.width=10, fig.height=7, dpi=300}
set.seed(130)
#umap, dim 1.40 and resolution 0.3
DimPlot(
  Norm_Post_QC_0.3,
  reduction = "umap.harmony",
  group.by = "seurat_clusters",
  label = TRUE,
  repel = TRUE
)

```

# QC on Clusters 0.3
```{r fig.width=10, fig.height=7, dpi=300}
VlnPlot(Norm_Post_QC_0.3, features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.rp", "scDblFinder.score"), ncol = 2, pt.size = 0) + theme(axis.title.x = element_blank())



```
# Canonical Markers 0.3
```{r }
canonical_markers <- c(
  "TNNT2", "ACTC1", "MYH6", "MYH7", "TNNI3", "RYR2", "PLN",
  "DCN", "LUM", "COL1A1", "COL1A2", "FBLN1", "COL3A1",
  "PECAM1", "CDH5", "KDR", "EMCN", "VWF", "EGFL7", "FLT1"
)

DotPlot(Norm_Post_QC_0.3, features = canonical_markers) + RotatedAxis()
```

# Export Seurat Object
```{r }
# Create the output folder if it does not already exist
dir.create(
  here::here("output"),
  showWarnings = FALSE,
  recursive = TRUE
)

# Save the complete initial clustered object
saveRDS(
  Norm_Post_QC_0.3,
  file = here::here(
    "output",
    "snRNAseq_initial_QC_and_Clustering.rds"
  )
)

#Check that it saved correctly
checkpoint_path <- here::here(
  "output",
  "snRNAseq_initial_clustered_before_iterative_QC.rds"
)

file.exists(checkpoint_path)
file.info(checkpoint_path)$size / 1024^3

```

# Conclusion
So we have problematic clusters: cluster number 3 (CM) has a high scDblFinder score. Cluster 9 has high percentages of mitochondrial genes and ribosomal proteins (CM), it also has relatively low features and counts. Cluster 7 has low features and counts and is positve for ECs and fibroblasts.
At this point I will save the Seurat Object, do further checks on the three problematic clusters, possibly remove some and redo QC. 

